neonstore: provenance-aware persistent storage for NEON data

### 2024-01-02 paper/paper.Rmd
wip

---
title: "neonstore: provenance-aware persistent storage for NEON data"
author:
  - name: "Carl Boettiger"
    affiliation: ucb
  - name: "R. Quinn Thomas"
    affiliation: vt
  - name: "Christine M. Laney"
    affiliation: neon
  - name: "Claire Lunch"
    affiliation: neon
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: neon
    address: "NEON"
  - code: vt
    address: "Virginia Tech"
abstract: |
  The National Ecological Observatory Network
journal: preprint
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: rticles::elsevier_article
---

@@

The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@nap_neon; @Balch2020; @Nagy2021].
NEON provides access to 181 (active and planned) ecological data products at 81 (permanent and relocatable) sites throughout the United States.
The `neonstore` R package seeks to provide provenance aware, high performance access and persistent storage of NEON data files. 
Here, we describe the design, use, and rationale for the package.
We illustrate a few example use cases, including downloading the 30 minute single-aspirated temperature data across all available NEON sites and dates, a collection of over 13 thousand individual files totalling nearly 2 gigabytes of data.
We demonstrate how `neonstore` facilitates summary calculations such as computing monthly means across all sites in a fraction of a second with minimal memory requirements.
The approaches taken in `neonstore` maximize reproducibility, facilitate collaboration including across computer languages, and help making analyses which leverage the full spatial and temporal scale of NEON easily available to researchers with modest computational resources and expertise.

Provenance and performance are two central aspects to the design of `neonstore`.
A focus on provenance means that analyses can always be traced back to individual NEON data files, and supports both workflows that require access to the most recent or updated data and workflows that require repeated access to persistent, stable data files.
A focus on performance reflects the potentially very large size of its aggregated data products, which can easily exceed available working memory (RAM) on most machines: as many unified tables of even single products over all available sites and dates can range from 10 GBs into the terrabyte range. 
Because enabling analyses that could span the full spatial and temporal range of NEON's 81 sites over a 30 year horizon is a foundational motivation for NEON research, it essential that researchers be able to easily work with data across this full span. 
However, doing so requires techniques that go beyond workflows which expect data to be re-downloaded (provenance) or stored in memory for analysis (performance).

To address these challenges, `neonstore` implements a high-performance, provenance aware workflow by drawing on a host of key technologies, including cryptographic hashes, a Lightning Memory Mapped Database (LMDB) for metadata management, a columnar-oriented integrated relational database (duckdb) for data management, and optional support for remote access through parquet files cloud-native object stores to facilitate sharing large databases generated by `neonstore`. 
These technologies are seamlessly integrated behind a user-friendly R interface which requires no expertise in the underlying backend technology beyond a familiarity with R and the "tidy data" model used in the popular R packages, `dplyr` and `tidyr`.

@@

`neonstore` seeks to help users download and manage NEON files locally in a way that lets analyses be quickly traced back to the raw data files.
`neonstore` is distinguished as much by what it does not do as by what it does.
All `neonstore` operations are intended to be transparent and intuitive, and could easily be duplicated in another computational language or command line interface.
Specifically, `neonstore` will never automatically "clean" data: opaque data transformations make output sensitive to the version and implementation details of the tool in question.
A user should expect the same file from a "download" operation regardless of the software or software version used to perform the download.
Functions which in read in the data should faithfully reproduce the raw data so that analyses can easily be compared to those using other software tools.
When functions perform tasks like guessing the data type (numeric, string, Date, etc) of a column, or adding additional columns to stack two otherwise matching tables in which one has recorded an additional variable, they should do so in a manner that is transparent, controllable, and consistent with the behaviour of common functions used in the language.

@@

## File Interface: `neon_index()` & `neon_read()`

@@

Consider the example of a much larger table collected by sensor data; say, single-aspirated temperature collected every 30 mintues across all sites:
Note that we exploit pattern matching to provide a short-hand to the table name.
We download and import the 30 minute sampled table. 
The initial download is not a fast operation - in our example, rate-limiting rules by NEON make the command take over 20 minutes to complete the API requests that would otherwise take only 3 minutes.
The download itself is determined more by network speed. 
In our example this takes about 1 hour to download over 13 thousand individual files.
Because NEON uses download links that expire after 1 hour, it may be necessary to repeat the download command after more than an hour expires,
(alternatively, divide the download into multiple batches by years or sites).
The `neon_download()` utility can pick up from where it left off.

@@

```{r include=FALSE}
temp %>% select(verticalPosition, horizontalPosition) %>% distinct()

index <- neon_index(product = "DP1.00002.001", table = "30min") 
info <- index$path %>% fs::file_info()
n_files <- length(index$path)
n_bytes <- sum(info$size)
```

**Cross-language support**: The `duckdb` database created by `neonstore` is not tied to R, but can easily be accessed using other languages supported by `duckdb`, including python, javascript, web assembly, and a stand-alone CLI interface [@duckdb].

## Network Database Interface

While most use cases will be covered by the interfaces discussed above, the networked database interface may be particularly useful for those wishing to scale analyses within or across teams.
The relational database interface still assumes that each user of the data individually downloads the NEON products they are using and imports that data into the local `duckdb` database on their machine.
While it is possible for multiple users to access the same database on a single shared server, it is common today for each researcher runs analyses on their own laptop, and also common for an individual researcher to work on their analyses on multiple machines (for instance, a laptop and a server).
Despite the streamlined tools available in `neonstore`, re-downloading and storing a large library of NEON files on many machines or users can be time-consuming and inefficient.
For example, downloading and importing the NEON data products required for the NEON Ecological Forecasting Challenge [@efi_neon] takes several hours and occupy over 100 GB of storage.
In such instances, network-based access provide a much more efficient mechanism for a team to interact with a shared `neonstore` database.
`neonstore` can take advantage of "cloud-native" interface mechanisms implemented in the Apache `arrow` [@arrow] library and the capabilities of the modern Apache `parquet` [@parquet] data serialization to provide shared access that does not require re-downloading or transferring the entire library to each machine, without the use of a central server. 
In this approach, one user takes the responsibility of running the `neon_download()` and `neon_store()` commands discussed above to create a database of containing the desired data products.
That user can then make this database available to other users through an S3 bucket [@aws.s3] or compatible system, such as the open-source software MINIO [@minio], using the function `neon_export_db()`.
Note that by only requiring a 'static' storage mechanism such as an S3 bucket, this approach is much cheaper and raises fewer security concerns than maintaining a central server running a database server such as MySQL.
Future development may enable access from other static storage services, including Azure cloud or GitHub (already possible in Python, see @fsspec).

Other team members can then access the remote database using the `neon_remote_db()` and `neon_remote()` functions, without ever having to download the data:
We pass the connection details to `neon_remote_db` (which can include authentication tokens if desired, see documentation for details.)
In this case, we will use a public (read-only) bucket:

```{r include=FALSE}
Sys.unsetenv("AWS_DEFAULT_REGION")
Sys.unsetenv("AWS_S3_ENDPOINT")
Sys.unsetenv("AWS_ACCESS_KEY_ID")
Sys.unsetenv("AWS_SECRET_ACCESS_KEY")
Sys.setenv("AWS_EC2_METADATA_DISABLED"="TRUE")
```

```{r cache=FALSE}
con <- neon_remote_db(host = "minio.thelio.carlboettiger.info",
                      bucket = "shared-data",
                      path = "neonstoredb")
```

Once a connection to the remote database is established, we can request individual tables using `neon_remote` to access the data using dplyr functions.
Like `neon_table()`, we must use `collect()` to draw the data into R if we wish to perform advanced R-specific operations that are not supported by the backend engine.

```{r  cache=FALSE}
remote_temp <- neon_remote("30min", product = "DP1.00002.001", db = con)
remote_temp %>% count(siteID) %>% collect()
```

Because this approach relies on operations over a network connection instead of a local disk, it will be most effective in contexts where high-speed networks are available, including local area networks and cloud computing platforms.
Alternately, Users may also download the `parquet` exported files directly and import them into a local `neonstore` instance (see `neon_import_db()`) to share neonstore snapshots without requiring independent downloads over the NEON API.
Using `neon_db_export()` to generate snapshots is also a reliable way to preserve and version your analyses for future that does not depend on access to the NEON API or even to `neonstore`.

**Cross-language access**: Once again, this approach should work equally well for data exchange with collaborators not using R.
Apache `arrow` can be used to directly access these remote parquet files from Python, Julia, Matlab, C++, Javascript, Rust, Go, and other popular languages in much the same way [@arrow]. 
This approach thus allows the `neonstore` user to facilitate efficient and scalable access to their NEON library not only across machines but also across languages.

<!--
In R, we refer to the `duckdb` database as a "remote" connection because the operations are handled outside of R, by `duckdb`.
Some database connections, like MySQL or Postgres, require the user to install a separate piece of software, often on an external server,
which we then access over the network connection (with appropriate security credentials), though it can also be run on the same local machine as the 'client' software.
This client-server approach can be very powerful, but requires more knowledge and resources to deploy.
In contrast, `duckdb` installs automatically as a stand-alone R package.
-->

>>>>

---
title: "neonstore: provenance-aware persistent storage for NEON data"
author:
  - name: "Carl Boettiger"
    affiliation: ucb
  - name: "R. Quinn Thomas"
    affiliation: vt
  - name: "Christine M. Laney"
    affiliation: neon
  - name: "Eric Sokol"
    affiliation: neon    
  - name: "Claire Lunch"
    affiliation: neon
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: neon
    address: "NEON"
  - code: vt
    address: "Virginia Tech"
abstract: |
  The National Ecological Observatory Network
journal: preprint
date: "`r Sys.Date()`"
bibliography: refs.bib
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: rticles::elsevier_article
---

@@

The National Ecological Observatory Network (NEON) represents an important resource and a major investment in data collection infrastructure needed to address many of the grand challenges facing environmental science [@nap_neon; @Balch2020; @Nagy2021].
NEON provides access to 181 (active and planned) ecological data products at 81 (permanent and relocatable) sites throughout the United States.
The `neonstore` R package seeks to provide provenance aware, high performance access and persistent storage of NEON data files. 
Here, we describe the design, use, and rationale for the package.
We illustrate a few example use cases, including working with larger-than-RAM data and cloud-native access patterns that avoid direct downloads.
The approaches taken in `neonstore` maximize reproducibility, facilitate collaboration including across computer languages, and help making analyses which leverage the full spatial and temporal scale of NEON easily available to researchers with modest computational resources and expertise.

Provenance and performance are two central aspects to the design of `neonstore`.
A focus on provenance means that analyses can always be traced back to individual NEON data files, and supports both workflows that require access to the most recent or updated data and workflows that require repeated access to persistent, stable data files.
A focus on performance reflects the potentially very large size of its aggregated data products, which can easily exceed available working memory (RAM) on most machines: as many unified tables of even single products over all available sites and dates can range from 10 GBs into the terrabyte range. 
Because enabling analyses that could span the full spatial and temporal range of NEON's 81 sites over a 30 year horizon is a fundamental motivation for NEON research, it essential that researchers be able to easily work with data across this full span. 
However, doing so requires techniques that go beyond workflows which expect data to be re-downloaded (provenance) or stored in memory for analysis (performance).

To address these challenges, `neonstore` implements a high-performance, provenance aware workflow by drawing on a host of key technologies, including cryptographic hashes, a Lightning Memory Mapped Database (LMDB) for metadata management, a columnar-oriented integrated relational database (duckdb) for data management, and  support for remote access through cloud-native object stores.
These technologies are seamlessly integrated behind a user-friendly R interface which requires no expertise in the underlying backend technology beyond a familiarity with R and the "tidy data" model used in the popular R packages, `dplyr` and `tidyr`.

@@

`neonstore` seeks to help users download and manage NEON files locally in a way that lets analyses be quickly traced back to the raw data files.
`neonstore` is distinguished as much by what it does not do as by what it does.
All `neonstore` operations are intended to be transparent and intuitive, and could easily be duplicated in another computational language or command line interface.
Specifically, `neonstore` will never automatically "clean" data: opaque data transformations make output sensitive to the version and implementation details of the tool in question.
A user should expect the same file from a "download" operation regardless of the software or software version used to perform the download.
Functions which in read in the data should faithfully reproduce the raw data so that analyses can easily be compared to those using other software tools.
When functions perform tasks like guessing the data type (numeric, string, Date, etc) of a column, or adding additional columns to stack two otherwise matching tables in which one has recorded an additional variable, they should do so in a manner that is transparent, controllable, and consistent with the behavior of common functions used in the language.

@@

## File-based Interface: `neon_index()` & `neon_read()`

@@

Consider the example of a much larger table collected by sensor data; say, single-aspirated temperature collected every 30 minutes across all sites.
We download and import the 30 minute sampled table
(ote that we exploit pattern matching to provide a short-hand to the table name).
The initial download is not a fast operation - in our example, rate-limiting rules by NEON make the command take over 20 minutes to complete the API requests that would otherwise take only 3 minutes.
The download itself is determined more by network speed. 
In our example this takes about 1 hour to download over 13 thousand individual files.

@@

```{r include=FALSE}
temp %>% select(verticalPosition, horizontalPosition) %>% distinct()
index <- neon_index(product = "DP1.00002.001", table = "30min") 
info <- index$path %>% fs::file_info()
n_files <- length(index$path)
n_bytes <- sum(info$size)
```

# Cloud-based access

Cloud-based access offers a simpler and sometimes faster alternative to the standard pattern of first downloading the data (with `neon_download()`), importing to the local database (with `neon_store()`) and then accessing from the database (with `neon_table()`).  In cloud-based access, we simply access data directly from the Google Cloud Storage platform without downloading:

```{r}
df <- neon_cloud(product = "DP1.00002.001", table = "30min") # Temp single aspirated
```

Unfortunately, at this time NEON still requires us to access the API to obtain a list of individual file URLs.  The API is quite slow, because that list of files requires many separate API requests (for each site and for each year-month in the request). In future this may become a single request and will thus be almost instantaneous.

`neon_cloud()` does not immediately download any of the actual data fields.  Instead, it merely reads the metadata (the table schema, i.e. column names) of the first CSV in the collection.  This approach is often referred to as "lazy evaluation" in software engineering design -- the function executes as quickly as possible by deferring any computation until it is absolutely necessary.  For instance, if we are going to work with only a few columns of the data, we can save computational effort by not downloading and parsing all those extra columns. Despite this strategy, our `df` object "feels" like a normal data.frame in most ways -- we can perform most of the usual operations as if the data were already both downloaded and read into memory.  Here we immediately perform the same calculation as before:

```{r}

temp_monthly <- df %>%
  mutate(month = month(startDateTime)) %>%
  group_by(siteID, month) %>%
  summarise(mean_temp = mean(tempSingleMean, na.rm=TRUE),
            max = max(tempSingleMean, na.rm=TRUE),
            min = min(tempSingleMean, na.rm=TRUE),
            .groups = "drop") %>% 
  collect()

```
